---
output: html_document
editor_options: 
chunk_output_type: console
---
---
title: "Reproducible Research - Project 1"
output: html_notebook

---
### Loading and preprocessing the data

```{r}
library("ggplot2")
library("dplyr")
library("timeDate")
data =  read.csv("activity.csv", sep = ",", header = TRUE)

data_clean <- na.omit(data)
```

### What is mean total number of steps taken per day?
1. Plotting the Histogram
  
```{r}

data_date <- data %>% 
              select(date, steps) %>%
              group_by(date)%>% 
              summarize(totalsteps= sum(steps))%>%
              na.omit()
```

```{r}
ggplot(data_date, aes(x=totalsteps)) + 
  geom_histogram(main="Total Steps in a day", 
                 color = 'blue', fill = 'white') + 
  labs( x = "Total Steps", y = "Frequency")
```
  
2. Calculate and report the mean and median total number of steps taken per day
  
```{r}
mean(data_date$totalsteps)
```
```{r}
median(data_date$totalsteps)
```

### What is the average daily activity pattern?

1.  Time series plot
```{r}
data_avg_date <- data_clean %>% 
              select(interval, steps)%>%
              group_by(interval) %>% 
              summarize(avgsteps= mean(steps))%>%
              na.omit()

ggplot(data_avg_date, 
       aes(x=interval, y=avgsteps)) +
  geom_line(size=1) + 
  labs( y = "Average number of steps", 
        y = "Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
 
```{r}
peak <- max(data_avg_date$avgsteps)
data_avg_date[((data_avg_date$avgsteps)==peak), ]
```
 
### Imputing missing values

1. Calculate and report the total number of missing values in the dataset
  
```{r}
sum(is.na(data))
```

2. Devise a strategy for filling in all of the missing values in the dataset
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
  
```{r}

replaceNA <- function(x) replace(x, is.na(x), floor(mean(x, na.rm = TRUE)))

data_new <- data %>%
            group_by(interval) %>%
            mutate(steps= replaceNA(steps))%>%
            ungroup()
head(data_new)  
```
  
4. Calculate and report the mean and median total number of steps taken per day
  
```{r}
data_date_new <- data_new %>% 
                group_by(date)%>% 
                summarize(totalsteps= sum(steps))%>%
                na.omit()%>%
                ungroup()
head(data_date_new)
```

```{r}
ggplot(data_date_new, 
       aes(x=totalsteps)) + 
  geom_histogram(main="Total Steps in a day", 
                 color = 'red', fill = 'white') + 
  labs( x = "Total Steps", 
        y = "Frequency")
```

```{r}
mean(data_date_new$totalsteps)
median(data_date_new$totalsteps)
```

##### Comments:
   + The mean remains the same even after rreplacing the NA with the average values
   + The overall data has shifted towards the mean after replacing the missing values with the average values 
   
### Are there differences in activity patterns between weekdays and weekends?

1.  Indicate whether a given date is a weekday or weekend day.
```{r}
data_new$day <- weekdays(as.Date(data_new$date))
data_new$wd <- ifelse(data_new$day=="Saturday" | data_new$day=="Sunday", "Weekend", "Weekday" )
head(data_new)
```

2. Panel Plot

```{r}
data_new_summ <- data_new %>% 
  group_by(wd, interval) %>% 
  summarise(avsteps = mean(steps))

data_new_summ %>% head()
ggplot(data_new_summ,
       aes(x=interval, y=avsteps, colour = wd)) +
  geom_line() +
  labs( y = "Average number of steps",
        y = "Interval") +
  facet_grid(rows = vars(wd))
```

